Spatial considerations in the analysis of conservation data: evaluating the presence  of \textit{Acer campestre} (field maple) – space is special

### 2016-10-05 paper_v1_word.Rmd
Add files via upload

---
title: 'Spatial considerations in the analysis of conservation data: evaluating the presence  of \textit{Acer campestre} (field maple) – space is special'
author: "Alexis Comber^1^, Steve Carver^1^, Paul Harris^2^, Carol Ximena Garzon-Lopez^3^, Duccio Rocchini^3^"
output: pdf_document
geometry: margin=3cm
---

@@

The paper tests the model suggested by Coudun et al (2006) but does this using a Geographically Weighted (GW) frameworks (Brunsdon et al, 1996) such as GW Regression (GWR), a spatial dependence model. Coudun et al (2006) found the presence of *Acer campestre* in France to be significantly related to two factors rainfall and evapotranspiration. Here Coudon’s analysis is extended to account for spatial autocorrelation in the variables which occurs when changes in properties of nearby features are found to be correlated and this contradicts the underlying assumption of independence in statistical analysis and inference. The result is spatial non-stationarity when the statistical pattern or relationship observed in one location differs from that in another. To explore these, this paper applies a GWR analysis to examine the spatial variation in the relationships between the predictor variables and Acer campestre presence to test for the presence of local, non-stationary relationships.

@@

## 2.1 Need for explicitly spatial methods 
\textbf{TBC - Harry?}

- Need: location is not just another variable - Key concept in spatial statistics / quant geog
- Review: GWR Not the first – similar ideas found in: Crop science (Fischer & Gosset 1935), Meteorology (Kolmogorov 1941; Gandin 1965); Mining (Krige 1951; Matheron 1963); Forestry (Matérn 1960); Theory (Yaglom 1955); Euripides: *Slight not what’s near, though aiming for what is far*
- Spatial statistics paradigms: Geostatistical models; Spatial regression models; Spatial point process models; Spatial ecology models; Spatial movement (trajectories) models; Spatial interaction models; Geographically weighted models etc: Point is that you choose one according to: (i) spatial process, (ii) spatial data type, (iii) inferential framework & (iv) research objectives
- Very brief overview of the GWR paradigm – detail is in sections later? And extensions

## 2.2 Review of spatial methods in ecology and conservation 
\textbf{TBC - Duccio?}
Understanding species distribution is a key concept in ecological research, in particular when dealing with biodiversity monitoring for conservation practices. Several papers have dealt with the importance of explicitly taking space into account when trying to model the distribution of a certain species, and, overall, its change in space and time.
The description of the distribution of biodiversity at different spatial and temporal scales
has long been the focus of ecology and biogeography. Reliable descriptions of species distributions
are fundamental for conservation and research purposes (Dormann, 2007). A number of statistical problems must be solved before spatial models can be built such as: the uncertainty related to the sampling of the modelled species (Rocchini et al., 2011), the correlation among predictors (Dormann et al....), the potential variability over space of the predictors being used (Rocchin et al., 2016, CAGEO), the spatio-temporal dependency of the predictors (Zuur and Ieno, 2016). 
Among these, spatial non stationairty and scale dependence ar expected to impact most of the ecological processes and relative patterns such as the distribution of biodiversity over space (Foody, 2004). 
In general , spatial ecological models assume stationarity of predictor variables....... 
...........
how non stationarity might affect results
...........

I will add refs.........

@@

The study by Coudun et al (2006) found the presence of *Acer campestre* to be significantly related to Autumn rainfall and actual Thornthwaite evapotranspiration. In this study rainfall, Potential evapotranspiration (PET) and wildness index data were used to construct a series of models for predicting the density of *Acer campestre* occurance. Data on rainfall were downloaded from the NERC Environmental Information Data Centre (Tanguy et al, 2015) which provides monthly 1km estimated rainfall data for each year. The average Autumn (3 month) rainfall was calculated for each 1km. Mean annual PET data were downloaded from the CGIAR Consortium for Spatial Information (Trabucco and Zomer, 2009) which provides global data at textbf\{0.0083 degrees, approximately 1km}. Finally, a wilder dataset was included in the model. This was to explore the degree to the presence of *Acer campestre* may be related to anthropogenic disturbance – anecdotally this species is frequently found in field margins and hedges. The Wilderness Quality Index data were generated for the whole of Europe at 1km resolution as described in Kuiters et al (2011) and can be considered as measure of non-anthropogenic activity. Each of these datasets were spatially aggregated over the OS 10m grid cells to generate mean values as shown in Figure 2.

@@

A multi-stage analysis was applied to model Acer campestre distributions. First, exploratory analyses were undertaken using a standard OLS regression to model distributions as an initial step. This identified significant predictor variables, under the assumption that relationships between predictor variables (rainfall, PET and wilderness) and species distributions are stationary (i.e. global). Then a GWR analysis was applied to examine the spatial variation in the relationships between the predictor variables and *Acer campestre* distributions (i.e. to test for the presence of local, non-stationary relationships). In overview, GW approaches use a moving window or kernel that passes through the study area. At each location being considered, data under the window are used to make a local calculation of some kind, such as a regression. The data are weighted by their distance to the kernel centre and in this way GW approaches construct a series of models at discrete locations in the study area. This is in contrast to global models, that consider all of the data (usually) in a single analysis of all data in the study area.

@@

```{r eval=T, echo=F, message=F, warning=F}
library(knitr)
kable(tab, booktabs = T, format = 'markdown') 
```

@@

```{r eval=T, echo=F, message=F, warning=F}
# convert the gw.model SDF to a Spatial Points Data Frame
library(GISTools)
m.gr2 = SpatialPointsDataFrame(m.g2$SDF,data.frame(m.g2$SDF), proj4string = os.proj)
```

@@

The potential for detrimental effects due to local collinearity has been ignored by nearly all of the GWR analyses reported in the literature, regardless of domain or subject. Collinearity occurs when one predictor variable has a strong positive or negative relationship with another, typically when it is less than -0.8 or greater than +0.8. Critically, collinearity may be absent when calculated globally (ie from all the data values), but may be present locally when a subset of the data is considered, as is the case with a GWR analysis. Table 3 shows the results of applying GWR collinearity diagnostics to the GWR model above. This generates local VDPs and local CNs at the same scale (i.e. using the same adaptive bandwidth of 21 data points). The results are summarised in Table 3 and mapped in Figure 5. These indicate a high degree of local collinearity, with CNs greater than 30 and VDPs greater than 0.5 using standard heuristics (Belsley et al 1980; O’Brien 2007), locally. These values suggest that the application of a locally-compensated ridge GWR is warranted.

```{r eval=T, echo=F, message=F, warning=F}
library(GWmodel)
```

@@

```{r eval=T, echo=F, message=F, warning=F}
### VIFs greater than 10 -> suggestive of collinearity
### CNs greater than 30 -> suggestive of collinearity
# summary(gw.col$SDF)
# VIFs & CNs for the variables considered
# summary(gw.col$SDF[,1:4])
tab3 <- gw.col$SDF@data[,1:4]
tab3 <- round(apply(tab3, 2, fivenum), 3)
rownames(tab3) <- c("Min", "1st Qu", "Median", "3rd Qu", "Max")
tab3 <- t(tab3)
rownames(tab3) <- c("Mean annual PET", "Mean Autumn rainfall","Mean Wilderness Quality Index", "Local CN")
```

\begin{center}
{Table 3. GWR collinearity measures, describing local CN and local VDPs for each predictor variable.}
\end{center}

@@

\newpage

@@

Akaike H (1973). Information Theory and an Extension of the Maximum Likelihood Princi-ple. In BN Petrov, F

Csaki (eds.), 2nd Symposium on Information Theory, pp. 267–281. Akademiai Kiado, Budapest.

>>>>

---
title: 'Spatial considerations in the analysis of conservation data: evaluating the presence  of \textit{Acer campestre} (field maple) – space is special'
author: "Alexis Comber^1^, Steve Carver^1^, Paul Harris^2^, Carol Ximena Garzon-Lopez^3^, Duccio Rocchini^3^"
output: word_document
geometry: margin=3cm
---

@@

In order to demenstrate the value of spatial dependence emethods, this  paper tests the model suggested by Coudun et al (2006) but does this using a Geographically Weighted (GW) framework (Brunsdon et al, 1996) in this case a GW Regression (GWR). Coudun et al (2006) found the presence of *Acer campestre* in France to be significantly related to two factors rainfall and evapotranspiration. Here Coudon’s analysis is extended to account for spatial autocorrelation in the variables which occurs when changes in properties of nearby features are found to be correlated and this contradicts the underlying assumption of independence in statistical analysis and inference. The result is spatial non-stationarity when the statistical pattern or relationship observed in one location differs from that in another. To explore these, this paper applies a GWR analysis to examine the spatial variation in the relationships between the predictor variables and Acer campestre presence to test for the presence of local, non-stationary relationships.

@@

Understanding species distribution and spatial dependencies is a key concept in ecological research (**REF needed**), in particular when dealing with biodiversity monitoring for conservation practices (**REF needed**). Several papers have dealt with the importance of explicitly taking space into account when trying to model the distribution of a certain species, and, overall, its change in space and time (**CITATIONS here**). The description of the distribution of biodiversity at different spatial and temporal scales has long been the focus of ecology and biogeography. Reliable descriptions of species distributions are fundamental for conservation and research purposes (Dormann, 2007). However, when modelling species spread, a number of statistical problems must be solved before spatial models can be built such as: the uncertainty related to the sampling of the modelled species (Rocchini et al., 2011), the correlation among predictors (**Dormann et al....**), the potential variability over space of the predictors being used (Rocchin et al., 2016, CAGEO), the spatio-temporal dependency of the predictors (Zuur and Ieno, 2016). Among these, spatial non stationairty and scale dependence are expected to impact most of the ecological processes and relative patterns such as the distribution of biodiversity over space (Foody, 2004).

The importance of spatial considerations has long been recognised in a number of disciplines: Fischer & Gosset (1935) in crop science; Kolmogorov (1941), Gandin (1965) in meteorology; Krige (1951) and  Matheron (1963) in mining; Matérn (1960) in forestry; theoretical developments by Moran (1950) and Yaglom (1955); Berry and Marble (1968) and Chorley and Haggett (1967) in geography; and Legendre and Legendre (1991) in ecology.

The major issue with modelling spatial data is the lack of independence of the spatial objects and the assumption of stationarity in the processes being modelled. These are particularly true with socio-economic data, but are also true of environmental and ecological data. Both concerns relate to Tobler's dictum (the first Law of Geography) that: '*everything is related to everything else, but near things are more related than distant things*' (Tobler 1970). However, in spite of numerous methodological advances in addressing such spatial modelling problems, policy related research is still routinely informed by non-spatial modelling practices, which are heroic at best, and ill-advised at worst. The lineage, from the late 1970s onwards, of such methodological advances can be loosely traced through spatial statistics texts from Journel and Huigbrechts (1978) to Cressie (1993) to Chilès and Delfiner (1999) to Cressie and Wilke (2011).

Focusing on regression models, the main drawback in assuming observations are independent is that any spatial dependencies turn up in the residuals, as they are not explicitly added to the model. For area-based data, ways to account for this include Besag's (1974) conditionally autoregressive model or Anselin's (1988) spatially autoregressive model.  In both cases, a criterion is needed for determining 'nearby' and is often supplied by the entries in a matrix of adjacencies. Wall (2004) notes that spatial covariance structures are implied by such matrices, which sometimes act as confounders in the analysis implying counterintuitive spatial processes. Vastly different covariance structures can be observed for different values of the autoregressive parameter. This is worrying for area-based modelling and is similar in nature to that found in Openshaw and Taylor's (1979) analysis of voting data in Iowa US, where different arrangements of the spatial units yielded correlations from -1 to +1 for the same pre-aggregated data. For point-based data, geostatistical regressions are commonly used, where spatial dependencies are modelled directly via the variogram, a function specifying a deterministic relationship between point pairs and their correlation. Recent advances have also seen geostatistical models applied in health studies where area-based data is common (Goovaerts 2009). These models are worthy in that they do not suffer from the pitfalls described by Wall (2004).

Spatial information can also be accounted for by expressing the regression's coefficients as functions of the spatial coordinates (Casetti 1982; Gorr and Oligschaeger 1994), permitting a model of relationship nonstationarity. A major advance in this area was advent of the GWR model (Brunsdon et al. 1996). Here spatial structures in the data are incorporated into the model through a continuous distance-decay weighting scheme. The spatial extent of the scheme is controlled through the bandwidth parameter, which is optimised to give the best model fit to the data. The resultant localised regression coefficients are mapped to display their spatial variation.  Brunsdon et al. (1998) and  Harris et al. (2015) provide some simple inferential mechanisms to determine whether the coefficients exhibit significant spatial variation; and other notable advances in the GWR method can be found in Nakaya et al. (2005), Wheeler (2007), Huang et al. (2010), Harris et al. (2011), Brunsdon et al. (2012) and Silva and Fotheringham (2015). However, a fully-coherent inferential structure of GWR remains to be developed.  It may be that, given GWR's kernel smoothing roots (e.g. see Cleveland 1979), the technique is best reserved for use as an exploratory and highly visual tool only, to act as a valuable precursor to a more sophisticated spatially-varying coefficient model, such as those proposed by Gelfand et al. (2003) or Assunção (2003).

The over-riding issue is that dependence and some form of nonstationarity are endemic in spatial data, and model forms should be made widely available which allow users to explore these structures both pre- and post-modelling. When spatial non-stationarity holds, the predictor and response variables is expected to change over space in an area.  Under spatial non-stationarity the relationship between the distribution of a certain species and the predictors shaping it may change over space. Thus any modelling activity  should also incorporate mechanisms to handle these characteristics of spatial data. This is an important challenge and this paper focuses on the modelling of spatial nonstationarity via the GWR model in order to demonstrate how to address process heterogeneity in relationships. The GWR model itself is formally described in Section 3, both in a basic form and an adapted form to deal with potential problems of localised collinearity in the predictor variables (Brunsdon et al. 2012; Gollini et al. 2015).

@@

The study by Coudun et al (2006) found the presence of *Acer campestre* to be significantly related to Autumn rainfall and actual Thornthwaite evapotranspiration. In this study rainfall, Potential evapotranspiration (PET) and wildness index data were used to construct a series of models for predicting the density of *Acer campestre* occurance. Data on rainfall were downloaded from the NERC Environmental Information Data Centre (Tanguy et al, 2015) which provides monthly 1km estimated rainfall data for each year. The average Autumn (3 month) rainfall was calculated for each 1km. Mean annual PET data were downloaded from the CGIAR Consortium for Spatial Information (Trabucco and Zomer, 2009) which provides global data at textbf\{0.0083 degrees, approximately 1km}. Finally, a wilder dataset was included in the model. This was to explore the degree to the presence of *Acer campestre* may be related to anthropogenic disturbance – anecdotally this species is frequently used as an ornamental tree and found in field margins and hedges. The Wilderness Quality Index data were generated for the whole of Europe at 1km resolution as described in Kuiters et al (2011)

**....more from Steve**

WQI can be considered as measure of non-anthropogenic activity. Each of these datasets were spatially aggregated over the OS 10m grid cells to generate mean values as shown in Figure 2.

@@

A multi-stage analysis was applied to model *Acer campestre* distributions. First, exploratory analyses were undertaken using a standard OLS regression to model distributions as an initial step. This identified significant predictor variables, under the assumption that relationships between predictor variables (rainfall, PET and wilderness) and species distributions are stationary (i.e. global). Then a GWR analysis was applied to examine the spatial variation in the relationships between the predictor variables and *Acer campestre* distributions (i.e. to test for the presence of local, non-stationary relationships). In overview, GW approaches use a moving window or kernel that passes through the study area. At each location being considered, data under the window are used to make a local calculation of some kind, such as a regression. The data are weighted by their distance to the kernel centre and in this way GW approaches construct a series of models at discrete locations in the study area. This is in contrast to global models, that consider all of the data (usually) in a single analysis of all data in the study area.

@@

```{r eval=T, fig.width=5, echo=F, message=F, warning=F}
library(knitr)
kable(tab, booktabs = T, format = 'markdown') 
```

@@

```{r eval=T, cache = T, echo=F, message=F, warning=F}
# convert the gw.model SDF to a Spatial Points Data Frame
library(GISTools)
m.gr2 = SpatialPointsDataFrame(m.g2$SDF,data.frame(m.g2$SDF), proj4string = os.proj)
```

@@

The potential for detrimental effects due to local collinearity has been ignored by nearly all of the GWR analyses reported in the literature, regardless of domain or subject. Collinearity occurs when one predictor variable has a strong positive or negative relationship with another, typically when it is less than -0.8 or greater than +0.8. Critically, collinearity may be absent when calculated globally (ie from all the data values), but may be present locally when a subset of the data is considered, as is the case with a GWR analysis. Table 3 shows the results of evaluating collinearity globally and locally using the GWR collinearity diagnostics tool included in the `GWmodel` R package.

Globally, the Variance Infation Factors (VIFs) are all less than 10, although 2 of the Variance Decomposition Proportions (VDPs) are greater than 0.5 and the Condition Number (CN) is greater than 30, using standard heuristics from Belsley et al (1980) and O’Brien (2007). Together these suggest the presence of variable collinearity when evaluated globally considering all data points together. Applying GWR collinearity diagnostics to the GWR model above generates local VIFs, VDPs and CNs at the same scale (i.e. using the same adaptive bandwidth of 21 data points). The results in Table 3 indicate a high degree of *local* collinearity in the CN, VIFs and VDPs. These values suggest that the application of a locally-compensated ridge GWR is warranted. The local collienarity measures are mapped in Figure 5.

```{r eval=T, echo=F, message=F, warning=F}
library(GWmodel)
library(car)
```

@@

```{r eval=T, cache = T, echo=F, message=F, warning=F}
gw.col.glob <- gwr.collin.diagno(d.a$tree~d.a$pet+d.a$rain+d.a$wqi, data = d.a, bw = nrow(d.a), kernel = "boxcar", adaptive = TRUE)
```

```{r eval=T, echo=F, message=F, warning=F}
### VIFs greater than 10 -> suggestive of collinearity
### CNs greater than 30 -> suggestive of collinearity
# summary(gw.col$SDF)
# VIFs & CNs for the variables considered
# summary(gw.col$SDF[,1:4])
tab3 <- gw.col$SDF@data[,c(1:4, 6:8)]
tab3 <- round(apply(tab3, 2, fivenum), 3)
rownames(tab3) <- c("Local Min", "Local 1st Qu", "Local Median", "Local 3rd Qu", "Local Max")
tab3 <- t(tab3)
row.i <- gw.col.glob$SDF@data[,c(1:4, 6:8)]
row.i <- round(apply(row.i, 2, fivenum), 3)[1,]
tab3 <- cbind(row.i, tab3)
colnames(tab3)[1] <- "Global"
rownames(tab3) <- c("VIF PET", "VIF Rainfall","VIF WQI", "CN", "VDP PET", "VDP Rainfall", "VDP WQI")
tab3 <- tab3[c(4,1:3,5:7),]
```

\begin{center}
{Table 3. Global and local collinearity measures: Condition Numbers with Variance Infaltion Factors (VIFs) and Variance Decomposition Proportions (VDPs) for each predictor variable.}
\end{center}

@@

@@

Akaike H (1973). Information Theory and an Extension of the Maximum Likelihood Princi-ple. In BN Petrov, F Csaki (eds.), 2nd Symposium on Information Theory, pp. 267–281. Akademiai Kiado, Budapest.
